Lessons learned from bugs in models of human history

### 2020-05-29 paper.tex
Small clarification in Fig 2 caption.
Add comparison to Martin et al violins in figure 2
Moved the ms model and into intro.

\documentclass{article}
\usepackage{fullpage}
\usepackage[round]{natbib}

\usepackage{listings}
\usepackage{url}

\usepackage{authblk}
\usepackage{graphicx}
\usepackage{color}

\lstset{language=Python}

@@

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is its Python API, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ require users to specify cryptic command
line options to describe demographic models.
Particularly for the large models we use today, these are not
intuitive or comprehensible for humans.
The Python API for \msprime, by contrast, allows the user to
state models in a more human-readable manner.
An example contrasting the different approaches to model specification
is provided in the Methods section.
%\sgcomment{Other software, such as SLiM,
%have developed advanced user interfaces and even dedicated programming languages
%to facilitate the specification of various modes of evolution.\cite{}
%}

Implementing multi-population models of demographic history is difficult and error-prone.
In this note we discuss two implementation errors that arose through
unfortunate design decisions in \msprime's demography API and which then found their way
into the scientific record. The first error has relatively mild effects on genetic diversity but was
used in many publications, while the second error was used only once but had a large impact on the
simulation results.
In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
discoveries and, more generally, best practices for implementing and simulating complex multi-population
demography.

@@

Even though the error has a limited effect on simulated data, the tutorial code has been
copied many times and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test a non-demographic inference
method, and the model was used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
% TODO ask Franz about this
Finally, \cite{pfaffelhuber2020choose}
used simulations of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

This long-standing error could have been prevented by better API design.
To model a population split currently in \msprime, a user must specify a
\texttt{MassMigration} event that moves lineages from one population to another
and then must also remember to turn off migration
between those populations at the same time.
The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
which more intuitively links the movement of lineages with appropriate changes in
migration rates at the time of the split.

@@

From a software perspective, this error was easy to make,
and could have been prevented by better API design.
The original \msprime\ API requires the user to pass
three separate parameters to specify a demographic model. For example,
\begin{lstlisting}[frame=single]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Martin et al. 2017
  migration_matrix=migration_matrix)
\end{lstlisting}
Note that the same three parameters must be passed to both the debugger
and the simulate function.
To help prevent such errors,
\msprime~ver.~1.0 will introduce a Demography class. The above
snippet would then be rewritten as
\begin{lstlisting}[frame=single]
demography = msprime.Demography(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{lstlisting}
This simple change to the interface makes it much less likely
that different models are passed to the debugger and simulator,
reducing the potential for error.

@@

For reference, the~\citet{gutenkunst2009inferring} demographic model, as written in \ms, is
\begin{lstlisting}[frame=single]
-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
-eg 0 2 116.010723 -eg 0 3 160.246047
-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
-ej 0.028985 3 2 -en 0.028985 2 0.287184
-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
-ej 0.197963 2 1 -en 0.303501 1 1
\end{lstlisting}
The equivalent model using \msprime\ is more human readable and provided in full at
\url{https://github.com/jeromekelleher/msprime-model-errors}. The demographic events
are given as
\begin{lstlisting}[frame=single]
demographic_events = [
    # CEU and CHB merge into B with rate changes at T_EU_AS
    msprime.MassMigration(
        time=T_EU_AS, source=2, destination=1, proportion=1.0),
    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
    msprime.PopulationParametersChange(
        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
    # Population B merges into YRI at T_B
    msprime.MassMigration(
        time=T_B, source=1, destination=0, proportion=1.0),
    msprime.MigrationRateChange(time=T_B, rate=0),
    # Size changes to N_A at T_AF
    msprime.PopulationParametersChange(
        time=T_AF, initial_size=N_A, population_id=0)
]
\end{lstlisting}
It is worth noting that the~\citet{gutenkunst2009inferring} model is relatively
simple, and models with many more populations and parameters are increasingly
common.

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
    demographic model.
    GWAS summary statistics were computed from 10,000 case and control subjects in the European
    population, and the distribution of inferred PRS were compared across the three simulated populations.
    Unlike the original study, we do not observe large differences in mean inferred PRS across
    the three populations (C, D), and while risk prediction in the African and East Asian populations is
    still reduced compared to the European population, the reduction in prediction accuracy is not as
    large as reported in the original study (E-G).
    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
}
\label{fig:prs}
\end{center}
\end{figure}

>>>>

\documentclass{article}
\usepackage[round]{natbib}
\usepackage{fullpage}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
\AfterEndEnvironment{minted}{\end{tcolorbox}}%
\usepackage{url}

@@

The \msprime\ coalescent
simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
is now widely used in genetics studies.
Much of its appeal is the large increase in efficiency over the classical
\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
samples of whole chromosomes for the first time. Another distinct advantage
of \msprime\ is its Python API, greatly
increasing the flexibility and ease of use over the standard approach of
text-based command line interfaces.
In particular, programs like \ms\ require users to specify cryptic command
line options to describe demographic models.
For example, the~\citet{gutenkunst2009inferring} demographic model
(that is the subject of this note), as written
in \ms\ syntax, is
\begin{minted}[]{tcsh}
-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
-eg 0 2 116.010723 -eg 0 3 160.246047
-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
-ej 0.028985 3 2 -en 0.028985 2 0.287184
-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
-ej 0.197963 2 1 -en 0.303501 1 1
\end{minted}
\noindent This model is relatively simple,
and models with many more populations and parameters are increasingly
common. Descriptions of such models in \ms\ syntax are not easy to comprehend.
The Python interface for \msprime, by contrast, allows the user to
state models in a more human-readable and programmatic manner,
and has many advantages over \ms's command line interface.

Even when using a high-level programming language like Python, however,
implementing multi-population models
of demographic history is difficult and error-prone.
In this note we discuss two implementation errors that arose through
unfortunate design decisions in \msprime's demography API and which then found their way
into the scientific record. The first error has relatively mild effects on genetic diversity but was
used in many publications, while the second error was used only once but had a large impact on the
simulation results.
In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
discoveries and, more generally, best practices for implementing and simulating complex multi-population
demography.

@@

Even though the error has a limited effect on simulated data, the tutorial code has been
copied many times and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test a non-demographic inference
method, and the model was used as an example of a complex population
history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
\cite{zhou2018popdemog} used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
Finally, \cite{pfaffelhuber2020choose}
used simulations of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

This long-standing error could have been prevented by better API design.
To model a population split currently in \msprime, a user must specify a
\texttt{MassMigration} event that moves lineages from one population to another
and then must also remember to turn off migration
between those populations at the same time.
The demographic events for the correct model are given as

\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
demographic_events = [
    # CEU and CHB merge into B with rate changes at T_EU_AS
    msprime.MassMigration(
        time=T_EU_AS, source=2, destination=1, proportion=1.0),
    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
    msprime.PopulationParametersChange(
        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
    # Population B merges into YRI at T_B
    msprime.MassMigration(
        time=T_B, source=1, destination=0, proportion=1.0),
    msprime.MigrationRateChange(time=T_B, rate=0), # NB This line was missing!!!
    # Size changes to N_A at T_AF
    msprime.PopulationParametersChange(
        time=T_AF, initial_size=N_A, population_id=0)
]
\end{minted}
Note that the migration rate change on 15 was missing in the incorrect model.
The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
which more intuitively links the movement of lineages with appropriate changes in
migration rates at the time of the split.

@@

From a software perspective, this error was easy to make
and could have been prevented by better API design.
The original \msprime\ API requires the user to pass
three separate parameters to specify a demographic model. For example,
\begin{minted}[highlightlines={8}]{python}
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Martin et al. 2017
  migration_matrix=migration_matrix)
\end{minted}
Note that the same three parameters must be passed to both the debugger
and the simulate function.
To help prevent such errors,
\msprime~ver.~1.0 will introduce a Demography class. The above
snippet would then be rewritten as
\begin{minted}[]{python}
demography = msprime.Demography(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{minted}
This simple change to the interface makes it much less likely
that different models are passed to the debugger and simulator,
reducing the potential for error.

@@

\section*{Acknowledgements}
\aprcomment{to fill in}
\begin{itemize}
\item Alicia - sharing data, comments?
\item Ilan - comments
\item Chris for comments?
\end{itemize}

@@

\begin{figure}[ht]
\begin{center}
\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
\caption{\textbf{The transferability of PRS under neutrality}.
    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
    events in the past, so continental populations were simulated as isolated with low levels of migration
    for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
    demographic model.
    GWAS summary statistics were computed from 10,000 case and control subjects in the European
    population, and the distribution of inferred PRS were compared across the three simulated populations.
    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
    Unlike the original study, we do not observe large differences in mean inferred PRS across
    the three populations (C, D), and while risk prediction in the African and East Asian populations is
    still reduced compared to the European population, the reduction in prediction accuracy is not as
    large as reported in the original study (E-G).
    Panel D shows a comparison of PRS distributions for a simulation replicate with 1,000 causal variants.
    For each population in the violin plots, the original correlations from~\citet{martin2017human}
    are shown on the left, and correlations using the correct model are shown on the right.
    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
}
\label{fig:prs}
\end{center}
\end{figure}
